---
title: "QUESTION 02: Data Pipeline"
output: html_document
date: "2024-11-26"
---

INTRODUCTION

"Sexual dimorphism" means that the two sexes of a species differ in external appearances or other features (Ralls & Mesnich., 2009). This includes differences in size, where one of the sexes are a lot larger than the other. Here, I investigate whether this sexual size dimorphism is apparent in the Gentoo penguin (Pygoscelis papua) using body mass data acquired from the "palmerpenguins" data set. If there was no size difference, we would expect the coefficient from a linear model analysis to be zero. If there was a size difference, we would epxect the coefficient from a linear model analysis to be significantly different form zero.


STATISTICAL METHODS

Loading libraries

```{r}
library(tidyverse)
library(palmerpenguins)
library(janitor)
library(here)
library(ragg)
library(svglite)
library(broom)
```


Loading functions

```{r}
#Cleaning function file
source(here("functions", "cleaning.R"))

#Function file to save figures as .png and .svg files
source(here("functions", "png_and_svg_functions.R"))

#Functions to produce exploratory figures 
source(here("functions", "exploratory_figure_functions.R"))

#Linear model figure function
source(here("functions", "linear_model_figure.R"))

#Diagnostic plot function
source(here("functions", "diagnostic_plot_function.R"))
```


Loading the data

```{r}
#Loading raw dataset
penguins_raw

#Saving raw dataset
write_csv(penguins_raw, here("data", "penguins_raw.csv"))
```


Cleaning and subsetting the data

```{r}
#Clean and subset the data. Filter by Gentoo species and subset species, body mass and sex
clean_gentoo_body_mass_sex <- penguins_clean %>%
  filter(species == "Gentoo") %>%   
  select(species, body_mass_g, sex) %>%  
  remove_NA() 

#visualise the data 
head(clean_gentoo_body_mass_sex)

#save cleaned data set
write_csv(clean_gentoo_body_mass_sex, here("data", "clean_gentoo_body_mass.csv"))
```


Creating an Exploratory Figure - Histogram

I first use a histogram to compare the differences in body mass between the sexes. From the figure below it appears that the modal body mass of males is about 250 grams heavier than females but the range of mass seems to be equal at about 1500 grams. Other than for the two male outliers and one female outlier the body mass data also appears to show little skew and looks normally distributed, which suggets that a linear model analysis would be a suitable statistical test to further compare the differences in body mass.

```{r}
#Histogram function
histogram <- overlapping_histogram_comparing_sexes(clean_gentoo_body_mass_sex, body_mass_g, sex, "Body Mass(g)", "Count")

#Visualising histogram 
histogram


#Saving histogram as .png 
save_as_png(clean_gentoo_body_mass_sex, 
            here("figures", "histogram.png"), 
            histogram,
            size = 20, res = 300, scaling = 2)


#Saving histogram as .svg
save_as_svg(clean_gentoo_body_mass_sex,
            here("figures", "histogram.svg"),
            histogram,
            size = 20, scaling = 2)
```


Defining the linear model 

```{r}
#Defining linear model
linear_model <- lm(body_mass_g ~ sex, data = clean_gentoo_body_mass_sex)
```


Checking for assumptions of linear model analysis 

Before performing a linear model analysis it is important to check that the Gentoo body mass data meets it's assumptions of: 1) independence 2) normality 3) homoscedasticity and 4) no influential outliers. I do not have access to the methods used to collect the body mass data but for the purpose of this analysis I have assumed that each data point is independent of each other. To then check for the other assumptions I have used a mulit-panel grid of diagnostic plots below. The Q-Q residuals plot shows that the data points closely follow the straight line suggesting that the residuals are normally distributed. Both the residuals vs fitted and the scale-location plot show a random scatter and a horizontal line indicating homoscedasticity. Finally, the data points in the residual vs leverage plot have a small Cook's distance so the outliers identified in the histogram are not having an influential effect. Since all of the assumptions for the linear model have been met, it is adequate to carry out the analysis.

```{r}
#Diagnostic plots 
#Displaying diagnostic plots in a multi-panel grid and adjusting margin size 
par(mfrow = c(2, 2), 
    mar = c(4, 4, 2, 2))  


#Diagnostic plot function
diagnostic_plot_function()


# Saving the diagnostic plot as a .png
save_diagnostic_plot_as_png(
  here("figures", "diagnostic_plot.png"),  
  diagnostic_plot_function,                    
  size = 20,                               
  res = 300,                              
  scaling = 2                              
)

# Saving the diagnostic plot as a .svg
save_diagnostic_plot_as_svg(
  here("figures", "diagnostic_plot.svg"),  
  diagnostic_plot_function,                    
  size = 20,                               
  scaling = 2                              
)
```


RESULTS

After carrying out a linear model analysis and summarising the results in a table, we can see that the male Gentoo penguins from the palmerpenguin data set are 805.09 grams heavier than the female Gentoo penguins. This difference in body mass is significant since the p < 0.05.

```{r}
#Linear model analysis 
summary(linear_model)
```


```{r}
#Linear model analysis summary table
#Firstly, converting model into a clean structure and extracting key details 
linear_model_summary <- tidy(linear_model)   

#Summary table - estimate, SE, and statistic rounded to three decimal points and p-values to 4 significant figures  
linear_model_summary_table <- linear_model_summary %>%
  select(term, estimate, std.error, statistic, p.value) %>%
  mutate(    
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = format(p.value, scientific = TRUE, digits = 4)) 

print(linear_model_summary_table)
```


```{r}
#Results figure function
linear_model_results_figure <-linear_model_figure(clean_gentoo_body_mass_sex, sex, body_mass_g, "Sex", "Body Mass (g)")

linear_model_results_figure

#Saving results figure as .png 
save_as_png(clean_gentoo_body_mass_sex, 
            here("figures", "results_figure.png"), 
            linear_model_results_figure,
            size = 20, res = 300, scaling = 2)


#Saving results figure as .svg
save_as_svg(clean_gentoo_body_mass_sex,
            here("figures", "results_figure.svg"),
            linear_model_results_figure,
            size = 20, scaling = 2)
```

DISCUSSION


CONCLUSION



QUESTION 3: Open Science
