---
title: "QUESTION 02: Data Pipeline"
output: html_document
date: "2024-11-26"
---

INTRODUCTION

"Sexual dimorphism" means that the two sexes of a species differ in external appearances or other features (Ralls & Mesnich., 2009). This includes differences in size, where one of the sexes are a lot larger than the other. Here, I investigate whether this sexual size dimorphism is apparent in the Gentoo penguin (Pygoscelis papua) using body mass data acquired from the "palmerpenguins" data set. If there was no size difference, we would expect the coefficient from a linear model analysis to be zero.


STATISTICAL METHODS

Loading libraries
```{r}
library(tidyverse)
library(palmerpenguins)
library(janitor)
library(here)
library(ragg)
library(svglite)
library(broom)
```


Loading functions
```{r}
#Cleaning function file
source(here("functions", "cleaning.R"))

#Function file to save figures as .png and .svg files
source(here("functions", "png_and_svg_functions.R"))

#Functions to produce exploratory figures 
source(here("functions", "exploratory_figure_functions.R"))

#Linear model figure function
source(here("functions", "linear_model_figure.R"))

#Diagnostic plot function
source(here("functions", "diagnostic_plot_function.R"))
```


Load the data
```{r}
#Loading raw dataset
penguins_raw

#Saving raw dataset
write_csv(penguins_raw, here("data", "penguins_raw.csv"))
```


Clean and subset the data
```{r}
#Clean and subset the data. Filter by Gentoo species and subset species, body mass and sex
clean_gentoo_body_mass_sex <- penguins_clean %>%
  filter(species == "Gentoo") %>%   
  select(species, body_mass_g, sex) %>%  
  remove_NA() 

#visualise the data 
head(clean_gentoo_body_mass_sex)

#save cleaned data set
write_csv(clean_gentoo_body_mass_sex, here("data", "clean_gentoo_body_mass.csv"))
```


Creating an Exploratory Figure - Histogram
```{r}
#Histogram function
histogram <- overlapping_histogram_comparing_sexes(clean_gentoo_body_mass_sex, body_mass_g, sex, "Body Mass(g)", "Count")

#Visualising histogram 
histogram


#Saving histogram as .png 
save_as_png(clean_gentoo_body_mass_sex, 
            here("figures", "histogram.png"), 
            histogram,
            size = 20, res = 300, scaling = 2)


#Saving histogram as .svg
save_as_svg(clean_gentoo_body_mass_sex,
            here("figures", "histogram.svg"),
            histogram,
            size = 20, scaling = 2)
```


Running a statistical test
```{r}
#Defining linear model
linear_model <- lm(body_mass_g ~ sex, data = clean_gentoo_body_mass_sex)
```


Checking for assumptions of linear model analysis 
```{r}
#Diagnostic plots 
#Displaying diagnostic plots in a multi-panel grid and adjusting margin size 
par(mfrow = c(2, 2), 
    mar = c(4, 4, 2, 2))  


#Diagnostic plot function
diagnostic_plot_function()


# Saving the diagnostic plot as a .png
save_diagnostic_plot_as_png(
  here("figures", "diagnostic_plot.png"),  
  diagnostic_plot_function,                    
  size = 20,                               
  res = 300,                              
  scaling = 2                              
)

# Saving the diagnostic plot as a .svg
save_diagnostic_plot_as_svg(
  here("figures", "diagnostic_plot.svg"),  
  diagnostic_plot_function,                    
  size = 20,                               
  scaling = 2                              
)
```


RESULTS
```{r}
#Linear model analysis summary table
#Firstly, converting model into a clean structure and extracting key details 
linear_model_summary <- tidy(linear_model)   

#Summary table - estimate, SE, and statistic rounded to three decimal points and p-values to 4 significant figures  
linear_model_summary_table <- linear_model_summary %>%
  select(term, estimate, std.error, statistic, p.value) %>%
  mutate(    
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 3),
    p.value = format(p.value, scientific = TRUE, digits = 4)) 

print(linear_model_summary_table)
```


```{r}
#Done
#Results figure function
linear_model_results_figure <-linear_model_figure(clean_gentoo_body_mass_sex, sex, body_mass_g, "Sex", "Body Mass (g)")

linear_model_results_figure

#Saving results figure as .png 
save_as_png(clean_gentoo_body_mass_sex, 
            here("figures", "results_figure.png"), 
            linear_model_results_figure,
            size = 20, res = 300, scaling = 2)


#Saving results figure as .svg
save_as_svg(clean_gentoo_body_mass_sex,
            here("figures", "results_figure.svg"),
            linear_model_results_figure,
            size = 20, scaling = 2)
```

DISCUSSION


CONCLUSION



QUESTION 3: Open Science
